Gametic Selection, Meiotic Drive, Sex Ratio Bias, and Transitions Between Sex Determination Systems

### 2017-02-10 LaTeX/Turnover.tex
Added some discussion/introduction, especially about sex ratios. Empirical predictions could still use expanding...

It has been suggested that sex ratio selection could be a particularly important force driving transitions between sex-determining systems \citep[Chapter 7]{Beukeboom:2014vb}. 
For example, flexible sex determination systems may be favoured in order to exploit local environmental conditions that are optimal for males or females, which creates locally biased sex ratios \citep{Charnov:1977tx,Werren:1984tl,Pen:2010kk}. 
In addition, feminizing mutations may invade when female biased sex ratios are favoured due to selection among demes \citep{Wilson:1981vm,Vuillleumier:2007bh}. 
In other situations, sex ratio selection may favour transitions in order to restore equal sex ratios. 
For example, \citet{Kozielska:2010vm} consider systems in which the ancestral sex chromosomes experience meiotic drive (e.g., where driving X or Y chromosomes are inherited disproportionately often), which causes sex ratios to become biased \citep{Hamilton:1967ts}. 
They find that new, unlinked sex-determining loci (masculinizing or feminizing mutations, i.e., neo-Y or neo-W loci) can then spread, restoring an even sex ratio.

@@

There are various ways in which a period of haploid selection could influence transitions between sex determination systems. 
Firstly, if we assume that haploid selection at any particular locus predominantly occurs in one sex (e.g., meiotic drive during spermatogenesis), then such loci experience a form of sex-specific selection. 
In this respect, we might expect that haploid selection might affect transitions between sex determination systems in a similar manner to sex-specific diploid selection \citep[as explored by][]{vanDoorn:2007eu,vanDoorn:2010hu}. 
That is, new masculizing mutations (neo-Y chromosomes) could be favoured via associations with alleles that are beneficial in the male haploid stage. 
However, sex ratios can also become biased by linkage between the sex-determining region and a locus that harbours genetic variation in haploid fitness. 
For example, there are several known cases of sex ratio bias caused by sex-linked meiotic drive alleles \citep[][, Chapter 3]{Burt:2006} or selection among X- and Y-bearing pollen \citep{Lloyd:1974tz,Conn:1981uw,Stehlik:2005ul,Stehlik:2006to,Field:2012fd,Field:2013cc}. 
It is not immediately clear how the spread of new sex determination systems would be influenced by the combination of sex ratio biases and associations between haploid selected loci and sex-determining regions.

Surprisingly, our models show that haploid selection influences the evolution of new sex determination systems in a way that is distinct from both diploid sex-specific selection and sex ratio selection. 
We find that new genetic sex determination systems are not affected by any sex ratio biases caused by associations between sex-determining regions and haploid selected loci. 
In addition, we find that associations that build up between an ancestral sex-determining locus and a haploid-selected locus can favour transitions between male and female heterogamety (e.g., a neo-W allele arising at a previously autosomal locus spreads in an ancestrally XY system), despite the fact that these ancestral associations were built up by selection. 
This does not occur in models that do not include haploid selection.

@@

%genetic system
We consider the transition between an ancestral and novel sex determination systems using a three locus model.
Locus \textbf{X} is the ancestral sex-determining region, with alleles $X$ and $Y$ (or $Z$ and $W$).
Locus \textbf{A} is a locus under selection, with alleles $A$ and $a$.
Locus \textbf{M} is a novel sex-determining region, at which the null allele ($M$) is initially fixed in the population such that sex of zygotes is determined by the genotype at the ancestral sex-determining region, \textbf{X} ($XX$ become females and $XY$ become males, or $ZW$ become females and $ZZ$ become males). 
To evaluate the evolution of new sex-determination systems, we consider the invasion, fixation, maintenance, and/or loss of novel sex-determining alleles ($m$) at the \textbf{M} locus. 
%We only do invasion analytically though (is this sentence still ok?)
We assume that the \textbf{M} locus is epistatically dominant over the \textbf{X} locus such that zygotes with at least one $m$ allele develop as females with probability $k$ and as males with probability $1-k$, regardless of the \textbf{X} locus genotype.
With $k=0$, the $m$ allele is a masculinizer (i.e., a neo-Y) and with $k=1$ the $m$ allele is a feminizer (i.e., a neo-W).
With intermediate $k$, the $m$ allele confers environmental sex determination such that zygotes develop as females in a proportion ($k$) of the environments they experience. 
Finally, we also analyze a model of maternally-controlled environmental sex-determination (ESD), where mothers with at least one $m$ allele produce daughters with probability $k$.

%life-cycle
In each generation, we census the genotype frequencies in male and female gametes/gametophytes (hereafter gametes) before gametic competition (see Sup.\ Mat. for recursion equations). 
First, competition occurs among male gametes (sperm/pollen competition) and among female gametes (egg/ovule competition) separately. 
Selection during gametic competition depends on the \textbf{A} locus genotype, relative fitnesses are given by $w_A^\Hermaphrodite$ and $w_a^\Hermaphrodite$ ($\Hermaphrodite \in \{\female,\male\}$; see table \ref{tab:fitnesstable}). %$w_{A}^\male$ and $w_{a}^\male$ for male gametes and $w_{A}^\female$ and $w_{a}^\female$ for female gametes, see table \ref{tab:fitnesstable}. 
We assume that all gametes compete for fertilization during gametic competition, which is not the case for monogamous mating systems where gametes from only one mating partner are present. 
Gametic competition in monogamous mating systems is equivalent to meiotic drive in our model, which only alters the frequency of gametes produced by heterozygotes. 
After gametic competition, random mating occurs between male and female gametes.
The resulting zygotes develop as males or females, depending on their genotypes at the \textbf{X} and \textbf{M} loci (and the \textbf{M} genotype of their mother in the case of maternal control) as described above.
Diploid males and females then experience selection, relative fitnesses are given by $w_{g}^{\male}$ in males and $w_{g}^{\female}$ in females, where $g$ is the diploid genotype at the \textbf{A} locus ($g \in \{AA, Aa, aa\}$).  
%Diploids compete with others of the same sex, where selection depends on the sex of the individual and its genotype at the \textbf{A} locus.
The next generation of gametes are then produced by meiosis, during which recombination and sex-specific meiotic drive can occur. 
Recombination occurs between loci \textbf{X} and \textbf{A} with probability $r$, between loci \textbf{A} and \textbf{M} with probability $R$, and between loci \textbf{X} and \textbf{M} with probability $\chi$.
Therefore, any order of the loci can be modelled with appropriate choices of $r$, $R$, and $\chi$ (see Table \ref{tab:chisubstitutions}). 
Males/females that are heterozygous at the \textbf{A} locus experience meiotic drive; $Aa$ heterozgotes of sex $\Hermaphrodite$ produce gametes bearing allele $A$ with probability $\alpha^\Hermaphrodite$. 
Thus, the \textbf{A} locus can experience sex-specific gametic competition, diploid selection and/or meiotic drive.

@@

\subsection*{Turnover between sex-determination systems} 
The evolution of a new sex determination system requires that a rare mutant allele, $m$, at the novel sex-determining locus  increases in frequency when rare. 
The spread of a rare mutant $m$ at the \textbf{M} locus is determined by the leading eigenvalue, $\lambda$, of the system described by the next generation frequency of eggs and sperm carrying the mutation, (\ref{eq:recursions}c), (\ref{eq:recursions}d), (\ref{eq:recursions}g), (\ref{eq:recursions}h), which is an eight equation system. %, evaluated at the above equilibrium.
%Thus, the system describing the change in frequency of the $m$ allele consists of eight recursion equations. 
Dominant neo-Y (when $k=0$) or neo-W alleles (when $k=1$) are only found in male diploids (neo-Y) or female diploids (neo-W) such that their growth rate ultimately depends only on the change in frequency of $m$-bearing gametes produced by males (for a neo-Y) or by females (for a neo-W). 
Furthermore, if the $m$ allele is fully dominant over the ancestral sex-determining system, phenotypes are not affected by the genotype at the ancestral sex-determining region (\textbf{X} locus). 
Therefore, the invasion of rare dominant neo-Y or neo-W alleles depends on only two equations and is therefore determined by the largest eigenvalue that solves the quadratic characteristic polynomial
$\lambda^2+ b \lambda + c =0$.
%\begin{equation}
%\lambda^2+ b \lambda + c =0
%\label{eq:charpoly_neoY}
%\end{equation}
%
%\noindent
%where $b$ is the average of the growth rates of the two haplotypes that carry the $m$ allele ($mA$ and $ma$), $b=(\lambda_{mA}+\lambda_{ma})/2$, and $c$ also involves the fitness of $m$ alleles when they recombine onto the other \textbf{A} background in a heterozygote, $c=\lambda_{mA}\lambda_{ma}+\rho_{mA}\rho_{ma}$ (see table \ref{tab:haplotype_growth}).
In this case $b= - (\lambda_{mA} + \lambda_{ma})$ and $c = \lambda_{mA} \lambda_{ma} -\rho_{mA} \rho_{ma}$, where $\lambda_{mi}-1$ is the growth rate of mutant haplotypes on background $i\in\{A,a\}$, accounting for loss due to recombination, and $\rho_{mi}-1$ is the rate of addition of mutant haplotypes onto background $i\in\{A,a\}$ due to recombination (see table \ref{tab:haplotype_growth}).
\textcolor{red}{check these interpretations as there are now two 2s in the lambdas of table 2. perhaps the 2 in the denominator is needed because the frequency of males sums to one as does the frequency of females? the two in the numerator is from heterozygotes (like Hardy-Weinberg)? should we absorb it into the drive or recombination term?}

@@

Table \ref{tab:haplotype_growth} illustrates a number of key points about the invasion of neo-Y and neo-W mutations. 
For a neo-Y, invasion depends on the relative lifetime fitness of $A$-bearing and $a$-bearing male gametes (i.e., sperm only).
The lifetime fitness of male gametes partly depends on the allele carried by the female gamete that they mate with (e.g., $A$ with probability $p_{X}^{\female} w_{A}^{\female}/\bar{w}_{H}^{\female}$).
Similarly, invasion of a neo-W depends on the relative lifetime fitness of $A$-bearing and $a$-bearing female gametes (i.e., eggs only). 
However, in the case of a neo-W, the allele carried by the male gamete that they mate with can come from either an X-bearing or a Y-bearing sperm (e.g., $A$ with probability $\bar{p}^{\male} w_{A}^{\male}/\bar{w}_{H}^{\male}$). 
In either case, the zygote will then develop as a female due to the presence of a neo-W. 
By contrast, females that do not carry the neo-W only result from matings with X-bearing sperm (e.g., matings with $A$-bearing sperm occur with probability $\bar{p}_{X}^{\male} w_{A}^{\male}/\bar{w}_{H}^{\male}$). 
If the \textbf{A} locus is initially linked to the ancestral sex-determining locus, \textbf{X}, the frequency of the $A$ allele among X- and Y-bearing sperm can differ (equation \ref{eq:freq_diffs}). 
Thus, eggs with and without a neo-W differ in the frequency of $A$ alleles they obtain from mating with male gametes.

We are particularly concerned with whether or not a rare neo-sex-determining allele increases in frequency, which occurs when the largest eigenvalue, $\lambda$, is greater than one. 
If the average change in frequency of the two haplotypes that carry the $m$ allele ($Am$ and $am$) is positive, invasion will always occur (i.e., if $\left\{(\lambda_{mA}-1)+ (\lambda_{ma}-1) \right\}/2 > 0$ then $\lambda > 1$). 
If neither haplotype increases in frequency ($\lambda_{mA}, \lambda_{ma} < 1$), the $m$ allele will not invade. 
Otherwise, the new sex-determining allele increases in frequency on one \textbf{A} background and declines on the other, and invasion requires
\begin{equation}\label{eq:neoYR}
r_{AM} \left[ 
\frac{p^\female_X w^\female_A w^\male_a (1-\alpha^\male)}{ \bar{w}^\female_H \bar{w}^\male_H (\lambda_{mA}-1)} + 
%p^\female_X \frac{w^\female_A}{\bar{w}^\female_H} \frac{w^\male_a(1-\alpha^\male)}{\bar{w}^\male_H} \frac{1}{\lambda_{mA}-1} + 
\frac{(1 - p^\female_X) w^\female_a w^\male_A \alpha^\male}{ \bar{w}^\female_H \bar{w}^\male_H (\lambda_{ma}-1)} 
\right] 
\frac{w^\male_{Aa}}{\bar{w}^\male} < 1,
\end{equation}

\noindent
for the neo-$W$. Equations \eqref{eq:neoYR} and \eqref{eq:neoWR} show that the new sex-determining allele, $m$, is expected to invade for any probability of recombination between loci \textbf{A} and \textbf{M}, $r_{AM}$, when the net flow of recombinants is from the less fit (smaller $\lambda_{mi}$) to the more fit \textbf{A} background (making the terms inside the square brackets in Equations \ref{eq:neoYR} and \ref{eq:neoWR} negative). \textcolor{red}{Q: is it definitely possible to have negative square brackets for a equilibria maintained by selection?}
%Assuming $g_A^* > 0$ and $g_a^* < 0$ without loss of generality, this is $-(1 - p^\female_X) w_{m,A} g_a^* < p^\female_X w_{m,a} g_A^*$.
When the net flow of recombinants is from the more fit to the less fit haplotype, the new sex-determining allele can still invade when the rate of recombination between it and the selected locus is small enough. \textcolor{red}{Q:Is it the case that sometimes the square brackets are positive and invasion occurs for $r_{AM}=1/2$? In which case it might be better to have slightly different phrasing here.}

%weak selection
We can explicitly determine the conditions under which invasion occurs if we assume that the $A$ allele reaches an equilibrium frequency under the ancestral sex-determination system before the neo-sex-determination system ($m$) arises. 
The equilibrium  frequency of $A$ on different ancestral backgrounds ($\hat{p}^\male_Y$, $\hat{p}^\male_X$, and $\hat{p}^\female_X$) is given by equations \eqref{eq:pAve} and \eqref{eq:freq_diffs} where we assume selection and meiotic drive are weak relative to recombination ($s^\Hermaphrodite$, $t^\Hermaphrodite$, $\alpha_{\Delta}^\Hermaphrodite$ of order $\epsilon$). 
Under weak selection, we denote the leading eigenvalue describing the invasion of a neo-Y ($k=0$) and a neo-W ($k=0$) into an ancestrally XY system by $\lambda_{Y',XY}$ and $\lambda_{W',XY}$, respectively, which are given by

@@

We find that neo-W alleles can invade for a large number of selective regimes. 
To clarify the parameter space under which $\lambda_{W',XY}>1$, we consider several special cases. 
Firstly, if the \textbf{A} locus is unlinked to the ancestral sex-determining region ($r=1/2$), a more closely linked neo-W ($R<1/2$) can always invade because $\left( \hat{p}^\male_Y-\hat{p}^\male_X \right)=0$ such that the second term in equation \eqref{eq:lambda_neoW} disappears and invasion depends only on the sign of $(r-R)$. 
Indeed, invasion typically occurs when the neo-W is more closely linked to the selected locus than the ancestral sex-determining region (Figure \ref{fig:Combination_Centimorgans}).
Secondly, we can simplify cases where invasion occurs despite $R>r$ using the special case where $R=1/2$ and $r<1/2$. 
In table \ref{tab:specialcases} we give the conditions where invasion occurs where we further assume that haploid selection only occurs during one phase in one sex (e.g., during male meiosis only) and dominance coefficients are equal in the two sexes, $h^\female=h^\male$. 
Where there is no gametic competition and meiotic drive in one sex only, an unlinked neo-W can invade as long as the same allele is favoured in male and female diploid selection ($s^\female s^\male>0$, see Figure \ref{fig:Combination_Centimorgans}B), which is 50\% of the parameter space. 
Where there is no meiotic drive and gametic competition occurs in one sex only, an unlinked neo-W can invade as long as the same allele is favoured in male and female diploid selection and there are sex differences in selection of one type ($s^\female(s^\male-s^\female)>0$, see Figure \ref{fig:Combination_Centimorgans}C,D), which is 25\% of the parameter space. 
These special cases indicate that neo-W invasion can occur for a relatively large fraction of parameter space, even if the neo-W is less tightly linked to the selected locus, $R>r$.

@@

%%write in general form without weak selection
%A rare, dominant neo-$Y$ ($k=0$) is always expected to invade the ancestral $XY$ system when the average growth rate of mutant haplotypes ($Am$ and $am$) is positive, $(g_A + g_a)/2 > 0$. %($g_i = where $\lambda_i$ are the two eigenvalues when $R=r=0$; the \textbf{X} locus does not affect the dynamics).
%When the growth rates of the mutant haplotypes without recombination ($R=0$) are negative, $g_i^* < 0 \; \forall \; i \in \{A,a\}$, where $g_i < g_i^*$, the neo-$Y$ does not invade.
%Otherwise, the mutant allele increases in frequency on one \textbf{A} background and declines on the other and neo-$Y$ invasion requires
%\begin{equation}
%R \left[ \frac{p^\female_X w_{m,a}}{g_A^*} + \frac{(1 - p^\female_X) w_{m,A}}{g_a^*} \right] w_{m,Aa} < \nu_m,
%\end{equation}
%where $w_{m,i}$ is the relative viability fitness of males depending on their haploid or diploid genotype at the \textbf{A} locus (with $i\in\{A,a,AA,Aa,aa\}$) and
%$\nu_m = 
%  p^\female_X p_{Ym} w^\male_{a} w^\male_{AA} +
%  p^\female_X (1-p_{Ym}) w^\male_{a} w^\male_{AA} +
%  (1-p^\female_X) p_{Ym} w^\male_{a} w^\male_{AA} +
%  (1-p^\female_X) (1-p_{Ym}) w^\male_a w^\male_{AA}$
%is the mean relative fitness of resident males.
%%and $g_j^*$ is the growth rate of $X^\male$ and $Ym$ haplotypes when on an $A$ ($j=A$)  or $a$ background ($j=a$).
%Neo-$Y$ invasion therefore occurs for any recombination rate, $R$, when the net flow of double recombinants is from the less fit to the more fit \textbf{A} background (making the term inside the square brackets negative).
%%Assuming $g_A^* > 0$ and $g_a^* < 0$ without loss of generality, this is $-(1 - p^\female_X) w_{m,A} g_a^* < p^\female_X w_{m,a} g_A^*$.
%When the net flow of double recombinants is from the more fit to the less fit haplotype, neo-$Y$ invasion can still occur when the rate of recombination is small enough.
%%weak selection
%Assuming weak selection, we can solve for the invasion fitness of the neo-$Y$ explicitly, giving
%\begin{equation}
%\lambda_{Y,XY} \approx 1 + V_A S_A (r - R)s^\female,
%\end{equation}

>>>>

It has been suggested that sex ratio selection could be a particularly important force driving transitions between sex-determining systems \citep[Chapter 7]{Beukeboom:2014vb}. 
The default mode of sex ratio selection is `Fisherian' sex ratio selection, which favours equal investment in male and female offspring (i.e., a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce)  \textcolor{red}{(Fisher)}. 
Given that the sex determination system can directly affect the sex ratio, we might expect Fisherian sex ratio selection to influence the spread of new sex determination systems. 
For example, \citet{Kozielska:2010vm} consider systems in which the ancestral sex chromosomes experience meiotic drive (e.g., where driving X or Y chromosomes are inherited disproportionately often), which causes sex ratios to become biased \citep{Hamilton:1967ts}. 
They find that new, unlinked sex-determining loci (masculinizing or feminizing mutations, i.e., neo-Y or neo-W loci) can then spread, which restoring an even sex ratio.

We note that non-Fisherian (`extraordinary') sex ratios can be favoured in some situations. 
For example, with local mate competition, where brothers compete for mates, a female biased sex ratio is favoured \citep{Hamilton:1967ts}. 
Thus, local mate competition could favour new sex determination systems because feminizing mutations bias the sex ratio towards females \citep{Wilson:1981vm,Vuillleumier:2007bh}. 
Another consideration is that environmental conditions (e.g., maternal condition, mate quality, age, or host size) can differentially affect the fitness of males versus females such that the optimal allocation to males/females depends on the environment (Trivers and Willard, 1973) \citep{Charnov:1977tx} (Charnov, 1982, West 2009, Chapters 6 and 7). 
In such cases, flexible sex determination systems may evolve in order to allow the primary sex ratio to be determined in a way that depends on the environment \citep{Charnov:1977tx,Werren:1984tl,Pen:2010kk}. 
In this study, we do not consider environmental condition dependence or local mate competition (reviewed in Charnov, 1982 and West 2009).

@@

There are various ways in which a period of haploid selection could influence transitions between sex determination systems. 
Firstly, if we assume that haploid selection at any particular locus predominantly occurs in one sex (e.g., meiotic drive during spermatogenesis), then such loci experience a form of sex-specific selection. 
In this respect, we might expect that haploid selection would affect transitions between sex determination systems in a similar manner to sex-specific diploid selection \citep[as explored by][]{vanDoorn:2007eu,vanDoorn:2010hu}. 
That is, new masculizing mutations (neo-Y chromosomes) could be favoured via associations with alleles that are beneficial in the male haploid stage. 
However, sex ratios can also become biased by linkage between the sex-determining region and a locus that harbours genetic variation in haploid fitness. 
For example, there are several known cases of sex ratio bias caused by sex-linked meiotic drive alleles \citep[][, Chapter 3]{Burt:2006} or selection among X- and Y-bearing pollen \citep{Lloyd:1974tz,Conn:1981uw,Stehlik:2005ul,Stehlik:2006to,Field:2012fd,Field:2013cc}. 
It is not immediately clear how the spread of new sex determination systems would be influenced by the combination of sex ratio biases and associations between haploid selected loci and sex-determining regions.

Surprisingly, our models show that haploid selection influences the evolution of new sex determination systems in a way that is distinct from both diploid sex-specific selection and Fisherian sex ratio selection. 
We find that the spread of new sex determination systems are independent of there being a zygotic sex ratio bias caused by associations between sex-determining regions and haploid selected loci. 
In addition, we find that associations that build up between an ancestral sex-determining locus and a haploid-selected locus can favour transitions between male and female heterogamety (e.g., a neo-W allele arising at a previously autosomal locus spreads in an ancestrally XY system), despite the fact that these ancestral associations were built up by selection. 
This does not occur in models that do not include haploid selection.

@@

%genetic system
We consider the transition between an ancestral and novel sex determination systems using a three locus model. 
Locus \textbf{X} is the ancestral sex-determining region, with alleles $X$ and $Y$ (or $Z$ and $W$).
Locus \textbf{A} is a locus under selection, with alleles $A$ and $a$.
Locus \textbf{M} is a novel sex-determining region, at which the null allele ($M$) is initially fixed in the population such that sex of zygotes is determined by the genotype at the ancestral sex-determining region, \textbf{X}; $XX$ genotypes become females and $XY$ become males (or $ZW$ become females and $ZZ$ become males). 
To evaluate the evolution of new sex-determination systems, we consider the invasion, fixation, maintenance, and/or loss of novel sex-determining alleles ($m$) at the \textbf{M} locus. 
%We only do invasion analytically though (is this sentence still ok?)
We assume that the \textbf{M} locus is epistatically dominant over the \textbf{X} locus such that zygotes with at least one $m$ allele develop as females with probability $k$ and as males with probability $1-k$, regardless of the \textbf{X} locus genotype.
With $k=0$, the $m$ allele is a masculinizer (i.e., a neo-Y) and with $k=1$ the $m$ allele is a feminizer (i.e., a neo-W).
With intermediate $k$, the $m$ allele confers environmental sex determination (ESD) such that zygotes develop as females in a proportion ($k$) of the environments they experience. 
Finally, we also analyze a model of maternally-controlled environmental sex-determination, where mothers with at least one $m$ allele produce daughters with probability $k$.

%life-cycle
In each generation, we census the genotype frequencies in male and female gametes/gametophytes (hereafter gametes) before gametic competition. 
A full description of our model, including recursion equations, is given in the Appendix. 
First, competition occurs among male gametes (sperm/pollen competition) and among female gametes (egg/ovule competition) separately. 
Selection during gametic competition depends on the \textbf{A} locus genotype, relative fitnesses are given by $w_A^\Hermaphrodite$ and $w_a^\Hermaphrodite$ ($\Hermaphrodite \in \{\female,\male\}$; see table \ref{tab:fitnesstable}). %$w_{A}^\male$ and $w_{a}^\male$ for male gametes and $w_{A}^\female$ and $w_{a}^\female$ for female gametes, see table \ref{tab:fitnesstable}. 
We assume that all gametes compete for fertilization during gametic competition, which is not the case for monogamous mating systems where gametes from only one mating partner are present. 
Gametic competition in monogamous mating systems is equivalent to meiotic drive in our model, which only alters the frequency of gametes produced by heterozygotes. 
After gametic competition, random mating occurs between male and female gametes.
The resulting zygotes develop as males or females, depending on their genotypes at the \textbf{X} and \textbf{M} loci (and the \textbf{M} genotype of their mother in the case of maternal control) as described above.
Diploid males and females then experience selection, relative fitnesses are given by $w_{g}^{\male}$ in males and $w_{g}^{\female}$ in females, where $g$ is the diploid genotype at the \textbf{A} locus ($g \in \{AA, Aa, aa\}$).  
%Diploids compete with others of the same sex, where selection depends on the sex of the individual and its genotype at the \textbf{A} locus.
The next generation of gametes are then produced by meiosis, during which recombination and sex-specific meiotic drive can occur. 
Recombination occurs between loci \textbf{X} and \textbf{A} with probability $r$, between loci \textbf{A} and \textbf{M} with probability $R$, and between loci \textbf{X} and \textbf{M} with probability $\chi$.
Therefore, any order of the loci can be modelled with appropriate choices of $r$, $R$, and $\chi$ (see Table \ref{tab:chisubstitutions}). 
Males/females that are heterozygous at the \textbf{A} locus experience meiotic drive; $Aa$ heterozgotes of sex $\Hermaphrodite$ produce gametes bearing allele $A$ with probability $\alpha^\Hermaphrodite$. 
Thus, the \textbf{A} locus can experience sex-specific gametic competition, diploid selection and/or meiotic drive.

@@

\subsection*{Turnover between sex-determination systems} 
The evolution of a new sex determination system requires that a rare mutant allele, $m$, at the novel sex-determining locus  increases in frequency when rare. 
The spread of a rare mutant $m$ at the \textbf{M} locus is determined by the leading eigenvalue, $\lambda$, of the system described by the next generation frequency of eggs and sperm carrying the mutation, (\ref{eq:recursions}c), (\ref{eq:recursions}d), (\ref{eq:recursions}g), (\ref{eq:recursions}h), which is an eight equation system. %, evaluated at the above equilibrium.
%Thus, the system describing the change in frequency of the $m$ allele consists of eight recursion equations. 
Dominant neo-Y (when $k=0$) or neo-W alleles (when $k=1$) are only found in male diploids (neo-Y) or female diploids (neo-W) such that their growth rate ultimately depends only on the change in frequency of $m$-bearing gametes produced by males (for a neo-Y) or by females (for a neo-W). 
Furthermore, if the $m$ allele is fully dominant over the ancestral sex-determining system, phenotypes are not affected by the genotype at the ancestral sex-determining region (\textbf{X} locus). 
Thus, the invasion of rare dominant neo-Y or neo-W alleles is determined by the largest eigenvalue that solves the quadratic characteristic polynomial
$\lambda^2+ b \lambda + c =0$.
%\begin{equation}
%\lambda^2+ b \lambda + c =0
%\label{eq:charpoly_neoY}
%\end{equation}
%
%\noindent
%where $b$ is the average of the growth rates of the two haplotypes that carry the $m$ allele ($mA$ and $ma$), $b=(\lambda_{mA}+\lambda_{ma})/2$, and $c$ also involves the fitness of $m$ alleles when they recombine onto the other \textbf{A} background in a heterozygote, $c=\lambda_{mA}\lambda_{ma}+\rho_{mA}\rho_{ma}$ (see table \ref{tab:haplotype_growth}).
In this case $b= - (\lambda_{mA} + \lambda_{ma})$ and $c = \lambda_{mA} \lambda_{ma} -\rho_{mA} \rho_{ma}$, where $\lambda_{mi}$ \textcolor{red}{I think we can call this growth rate, rather than reserving "growth rate" for $\lambda-1$. e.g., in ecological models, can't we call both R (in discrete time) and r (in continuous time) "growth rates"?} is the growth rate of mutant haplotypes on background $i\in\{A,a\}$, accounting for loss due to recombination, and $\rho_{mi}$ is the rate of addition of mutant haplotypes onto background $i\in\{A,a\}$ due to recombination (see table \ref{tab:haplotype_growth}).
\textcolor{red}{check these interpretations. the $\lambda$s are certainly related to true growth rates because we get them by looking at the case where $R=0$ so that you only have the haplotypes growing/declining. }
% as there are now two 2s in the lambdas of table 2. perhaps the 2 in the denominator is needed because the frequency of males sums to one as does the frequency of females?}

@@

Table \ref{tab:haplotype_growth} illustrates a number of key points about the invasion of neo-Y and neo-W mutations. 
For a neo-Y, invasion depends on the relative lifetime fitness of $A$-bearing and $a$-bearing male gametes (i.e., sperm only).
The lifetime fitness of male gametes partly depends on the allele carried by the female gamete that they mate with (e.g., $A$ with probability $p_{X}^{\female} w_{A}^{\female}/\bar{w}_{H}^{\female}$).
Similarly, invasion of a neo-W depends on the relative lifetime fitness of $A$-bearing and $a$-bearing female gametes (i.e., eggs only). 
However, in the case of a neo-W, the allele carried by the male gamete that they mate with can come from either an X-bearing or a Y-bearing sperm (e.g., $A$ with probability $\bar{p}^{\male} w_{A}^{\male}/\bar{w}_{H}^{\male}$, where $\bar{p}^{\male}=p_{Y}^{\male}q+p_{X}^{\male}(1-q)$). 
In either case, the zygote will then develop as a female due to the presence of a neo-W. 
By contrast, females that do not carry the neo-W only result from matings with X-bearing sperm (e.g., matings with $A$-bearing sperm that result in females occur with probability $(1-q)\bar{p}_{X}^{\male} w_{A}^{\male}/\bar{w}_{H}^{\male}$). 
If the \textbf{A} locus is initially linked to the ancestral sex-determining locus, \textbf{X}, the frequency of the $A$ allele among X- and Y-bearing sperm can differ (equation \ref{eq:freq_diffs}). 
Thus, eggs with and without a neo-W differ in the frequency of $A$ alleles they obtain from mating with male gametes.

%weak selection
We are particularly concerned with whether or not a rare neo-sex-determining allele increases in frequency, which occurs when the largest eigenvalue, $\lambda$, is greater than one. 
In the Appendix, we derive these conditions without assuming that selection is weak relative to recombination. 
Here, we explicitly determine the conditions under which invasion occurs by assuming that the $A$ allele reaches an equilibrium frequency under the ancestral sex-determination system before the neo-sex-determination system ($m$) arises. 
The equilibrium  frequency of $A$ on different ancestral backgrounds ($\hat{p}^\male_Y$, $\hat{p}^\male_X$, and $\hat{p}^\female_X$) is given by equations \eqref{eq:pAve} and \eqref{eq:freq_diffs} where we assume selection and meiotic drive are weak relative to recombination ($s^\Hermaphrodite$, $t^\Hermaphrodite$, $\alpha_{\Delta}^\Hermaphrodite$ of order $\epsilon$). 
Under weak selection, we denote the leading eigenvalue describing the invasion of a neo-Y ($k=0$) and a neo-W ($k=1$) into an ancestrally XY system by $\lambda_{Y',XY}$ and $\lambda_{W',XY}$, respectively, which are given by

@@

We find that neo-W alleles can invade for a large number of selective regimes. 
To clarify the parameter space under which $\lambda_{W',XY}>1$, we consider several special cases. 
Firstly, if the \textbf{A} locus is unlinked to the ancestral sex-determining region ($r=1/2$), a more closely linked neo-W ($R<1/2$) can always invade because $\left( \hat{p}^\male_Y-\hat{p}^\male_X \right)=0$ such that the second term in equation \eqref{eq:lambda_neoW} disappears and invasion depends only on the sign of $(r-R)$. 
Indeed, invasion typically occurs when the neo-W is more closely linked to the selected locus than the ancestral sex-determining region (Figure \ref{fig:Combination_Centimorgans}).
Secondly, we can simplify cases where invasion occurs despite $R>r$ using the special case where $R=1/2$ and $r<1/2$. 
In table \ref{tab:specialcases} we give the conditions where invasion occurs where we further assume that haploid selection only occurs during one phase in one sex (e.g., during male meiosis only) and dominance coefficients are equal in the two sexes, $h^\female=h^\male$. 
Where there is no gametic competition and meiotic drive in one sex only, an unlinked neo-W can invade as long as the same allele is favoured in male and female diploid selection ($s^\female s^\male>0$, see Figure \ref{fig:Combination_Centimorgans}B), which is 50\% of the parameter space. 
Where there is no meiotic drive and gametic competition occurs in one sex only, an unlinked neo-W can invade as long as the same allele is favoured in male and female diploid selection and there are sex differences in selection of one type (e.g., $s^\female(s^\male-s^\female)>0$, see Figure \ref{fig:Combination_Centimorgans}C,D), which is 25\% of the parameter space. 
These special cases indicate that neo-W invasion can occur for a relatively large fraction of parameter space, even if the neo-W is less tightly linked to the selected locus, $R>r$.

@@

We next consider the case where the new sex-determining mutation, $m$, causes sex to be determined by environmental conditions (environmental sex determiner, ESD). 
We assume that individuals carrying the $m$ allele develop as females in a fraction, $k$, of the environments they experience. 
The spread of these mutations is given by

\begin{equation}
\begin{split}
\lambda_{ESD',XY} =& 1 + (1-2k)^2V_{A}{S_{A}}^2\frac{r-R}{r R} \\
&+\frac{k(\hat{p}^\male_Y-\hat{p}^\male_X)}{2}\left(k\left(2\alpha_{\Delta}^\male-2\alpha_{\Delta}^\female+t^\male-t^\female \right) -4(1-k)S_{A}\right)+O\left(\epsilon^3\right),
\end{split}
\label{eq:lambda_ESD_k}
\end{equation}

\noindent
which reduces to $\lambda_{Y',XY}$ when $k=0$ and $\lambda_{W',XY}$ when $k=1$.
